Automatic Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Handlebar Perturbations\\Version 2.dev

### 2024-09-26 main.tex
Added acknowledgement to Allen Downey.
Added coda and data availability.
Address Marten's feedback.

\abstract{
  Uncontrolled bicycles are generally unstable at low speeds. We add an
  automatically controlled steering motor to a consumer electric bicycle that
  has a stabilizing effect down to about 3~\si{\kph}. We hypothesize that a
  stabilized bicycle will reduce the probability of falling. To test the
  motor's possible assistance during falls, we apply varying magnitude external
  handlebar perturbations to twenty-six participants who rode on a treadmill with
  the balance assist system activated and deactivated. The probability of not
  recovering from a handlebar perturbation significantly decreases when the
  balance assist is activated at a travel speed of 6~\si{\kph}. This positive
  effect is most prominent at and around the individual riders' perturbation
  resistance threshold. We conclude that use of a balance assist system in real
  world bicycling would reduce the number of falls that occur near riders'
  control authority limits.
}

\section{Introduction}
%
Single-actor bicycle crashes are associated with a surprisingly large
percentage of reported serious injuries~\citep{Wegman2024}. At low speeds, from
start up to typical cruising speeds, bicycles themselves are not self-stable
and can be challenging for the rider to balance. Low speed crashes may be
reduced if the bicycle was self-stable at these speeds by relieving the rider
of some of their necessary control activity. Bicycles can be mechanically
modified to lower the speeds at which they are self-stable~\citep{Astrom2005}
and since the 1980s it has been known that applying a motor actuated steering
torque proportional to the vehicle's roll angular rate~\footnote{Roll angular
rate is defined herein as per the SAE J670 vehicle standards.} can stabilize a
single-track vehicle down to very low speeds. If automatic control of steering
can stabilize a bicycle, it may reduce the control required from the rider to
successfully manage balancing tasks much like the natural high speed
self-stability already does. We have developed a balance-assisting bicycle,
Figure~\ref{fig:balance-assist-bicycle}, based on these principles and
hypothesize that it helps the rider in situations in which they are likely to
fall.
%
\begin{figure}
  \centering
  \includegraphics[width=80mm]{figures/balance-assist-bicycle.jpg}
  \caption{Balance assist bicycle prototype with electric motor in the steering
    column and data acquisition and control electronics mounted in the rear
    rack.}
  \label{fig:balance-assist-bicycle}
\end{figure}

@@

\subsection{Balance Assist Control}
%
We use a forward speed \(v\) gain scheduled proportional roll angular rate
feedback controller to stabilize the bicycle. In the speed range tested, the
commanded steer torque \(T^\textrm{m}_\delta\) from the steer motor follows the
control law
%
\begin{align}
  T^\textrm{m}_\delta = -k_{\dot{\phi}}\dot{\phi} = g(v_{stable} - v)\dot{\phi}
  \label{eq:implemented-controller}
\end{align}
%
where \(v_{\textrm{stable}} = 4.7~\si{\meter\per\second}\) is approximately the
average stable speed predicted from the open loop bicycle rigid-rider dynamics.
We use \(g=8\) and \(g=10\) to tune the gain magnitude during the experiments.
Scaling the proportional feedback gain linearly with respect to speed
stabilizes the normally unstable weave mode of the bicycle down to
2.9~\si{\kilo\meter\per\hour} for the riderless bicycle and
4.6~\si{\kilo\meter\per\hour} for the ridden bicycle as shown in
Figure~\ref{fig:eigenvalues}.
%
\begin{figure}
  \centering
  \includegraphics[width=160mm]{figures/balance-assist-eig-vs-speeds.png}
  \caption{Uncontrolled (upper row) and controlled with \(g=10\) (lower row)
    root locus of eigenvalue components (real: solid, imaginary: dashed)
    plotted versus speed for the bicycle without the inertial effect of a rigid
    rider (left) and with a rigid rider (right). Vertical dotted and
    dotted-dashed lines indicate the two speeds we perform experiments at. Grey
    shaded region is the linear stable speed range. Geometric and inertial
    parameters for these plots were estimated with the methods presented in
    \citet{Moore2012} and software packages BicycleParameters
    1.1.1~\citep{Moore2024} and Yeadon 1.5.0~\citep{Dembia2015} and are shown
    in Appendix~\ref{sec:physical-parameters}.}
  \label{fig:eigenvalues}
\end{figure}

@@

\section*{Acknowledgements}
%
This study follows and draws from experimental and analysis methods originally
developed in unpublished research by Marco Reijne. The authors acknowledge
Felix Dauer, David Gabriel, Sierd Heida, Oliver Maier, Maarten Pelgrim, Marco
Reijne, and Arend Schwab for contributions to the development of the balance
assist bicycle. We acknowledge Shannon van de Velde and Marco Reijne's
development of the bicycle perturbation system. We thank all of the assistants
that helped run the experiments and the participants for their time an
enthusiasm. We thank Peter Stahlecker for valuable feedback in early revisions.

@@

>>>>

\abstract{
  Uncontrolled bicycles are generally unstable at low speeds. We add an
  automatically controlled steering motor to a consumer electric bicycle that
  has a stabilizing effect down to about 3~\si{\kph}. We hypothesize that a
  stabilized bicycle will reduce the probability of falling. To test the
  motor's possible assistance during falls, we apply varying magnitude external
  handlebar perturbations to twenty-six participants who rode on a treadmill
  with the balance assist system activated and deactivated. The probability of
  recovering from a handlebar perturbation significantly increases when the
  balance assist is activated at a travel speed of 6~\si{\kph}. This positive
  effect is most prominent at and around the individual riders' perturbation
  resistance threshold. We conclude that use of a balance assist system in real
  world bicycling would reduce the number of falls that occur near riders'
  control authority limits.
}

\section{Introduction}
%
Single-actor bicycle crashes are associated with a surprisingly large
percentage of reported serious injuries~\citep{Wegman2024}. At low speeds, from
start up to typical cruising speeds, bicycles are not self-stable and can be
challenging for the rider to balance. Low speed crashes may be reduced if the
bicycle was self-stable at these speeds by relieving the rider of some of their
necessary control activity. Bicycles can be mechanically modified to lower the
speeds at which they are self-stable~\citep{Astrom2005} and since the 1980s it
has been known that applying a motor actuated steering torque proportional to
the vehicle's roll angular rate~\footnote{Roll angular rate is defined herein
as per the SAE J670 vehicle standards.} can stabilize a single-track vehicle
down to very low speeds. If automatic control of steering can stabilize a
bicycle, it may reduce the control required from the rider to successfully
manage balancing tasks much like the natural high speed self-stability already
does. We have developed a balance-assisting bicycle,
Figure~\ref{fig:balance-assist-bicycle}, based on these principles and
hypothesize that it helps the rider in situations in which they are likely to
fall.
%
\begin{figure}
  \centering
  \includegraphics[width=80mm]{figures/balance-assist-bicycle.jpg}
  \caption{Balance assist bicycle prototype with electric motor in the steering
    column and data acquisition and control electronics mounted in the rear
    rack.}
  \label{fig:balance-assist-bicycle}
\end{figure}

@@

\subsection{Balance Assist Control}
%
We use a forward speed \(v\) gain scheduled proportional roll angular rate
feedback controller to stabilize the bicycle. In the speed range tested, the
commanded steer torque \(T^\textrm{m}_\delta\) from the steer motor follows the
control law
%
\begin{align}
  T^\textrm{m}_\delta = -k_{\dot{\phi}}\dot{\phi} = g(v_{stable} - v)\dot{\phi}
  \label{eq:implemented-controller}
\end{align}
%
where \(v_{\textrm{stable}} = 4.7~\si{\meter\per\second}\) is approximately the
average stable speed predicted from the open loop bicycle rigid-rider dynamics.
We use \(g=8\) (low) and \(g=10\) (high) to tune the gain magnitude during the
experiments.  Scaling the proportional feedback gain linearly with respect to
speed stabilizes the normally unstable weave mode of the bicycle down to
2.9~\si{\kilo\meter\per\hour} for the riderless bicycle and
4.6~\si{\kilo\meter\per\hour} for the ridden bicycle as shown in
Figure~\ref{fig:eigenvalues}.
%
\begin{figure}
  \centering
  \includegraphics[width=160mm]{figures/balance-assist-eig-vs-speeds.png}
  \caption{Uncontrolled (upper row) and controlled with \(g=10\) (lower row)
    root locus of eigenvalue components (real: solid, imaginary: dashed)
    plotted versus speed for the bicycle without the inertial effect of a rigid
    rider (left) and with a rigid rider (right). Vertical dotted and
    dotted-dashed lines indicate the two speeds we perform experiments at. Grey
    shaded region is the linear stable speed range. Geometric and inertial
    parameters for these plots were estimated with the methods presented in
    \citet{Moore2012} and software packages BicycleParameters
    1.1.1~\citep{Moore2024} and Yeadon 1.5.0~\citep{Dembia2015} and are shown
    in Appendix~\ref{sec:physical-parameters}.}
  \label{fig:eigenvalues}
\end{figure}

@@

\section*{Acknowledgements}
%
This study follows and draws from experimental and analysis methods originally
developed in unpublished research by Marco Reijne. The authors acknowledge
Felix Dauer, David Gabriel, Sierd Heida, Oliver Maier, Maarten Pelgrim, Marco
Reijne, and Arend Schwab for contributions to the development of the balance
assist bicycle. We acknowledge Shannon van de Velde and Marco Reijne's
development of the bicycle perturbation system. Allen Downey provided useful
insight in the statistical condsiderations. We thank all of the assistants that
helped run the experiments and the participants for their time an enthusiasm.
We thank Peter Stahlecker for valuable feedback in early revisions.

@@

\section*{Code and Data Availability}
%
The data and code to reproduce the paper are available at:

\noindent
\url{https://github.com/mechmotum/fall-probability-paper}
